---
title: "lec16.Rmd"
author: "Adam Kapelner"
date: "April 28, 2020"
output: html_document
---

#Using AUC to Compare Probabilistic Classification Models

What would the effect be of less information on the same traing set size? Imagine we didn't know the features: occupation, education, education_num, relationship, marital_status. How would we do relative to the above? Worse!

```{r}
pacman::p_load(data.table, tidyverse, magrittr)
if (!pacman::p_isinstalled(ucidata)){
  pacman::p_load_gh("coatless/ucidata")
} else {
  pacman::p_load(ucidata)
}

data(adult)
adult = na.omit(adult) #kill any observations with missingness

set.seed(1)
train_size = 5000
train_indices = sample(1 : nrow(adult), train_size)
adult_train = adult[train_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL

test_size = 5000
test_indices = sample(setdiff(1 : nrow(adult), train_indices), test_size)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL

logistic_mod = glm(income ~ ., adult_train, family = "binomial")
p_hats_test = predict(logistic_mod, adult_test, type = "response")

performance_metrics_oos_full_mod = compute_metrics_prob_classifier(p_hats_test, y_test) %>% data.table

logistic_mod_less_data = glm(income ~ . - occupation - education - education_num - relationship - marital_status, adult_train, family = "binomial")
p_hats_test = predict(logistic_mod_less_data, adult_test, type = "response")

performance_metrics_oos_reduced_mod = compute_metrics_prob_classifier(p_hats_test, y_test) %>% data.table


ggplot(rbind(
  performance_metrics_oos_full_mod[, model := "full"],
  performance_metrics_oos_reduced_mod[, model := "reduced"]
)) +
  geom_point(aes(x = FPR, y = recall, shape = model, col = p_th), size = 1) +
  geom_abline(intercept = 0, slope = 1) + 
  coord_fixed() + xlim(0, 1) + ylim(0, 1) + 
  scale_colour_gradientn(colours = rainbow(5))
```

and we can see clearly that the AUC is worse. This means that the full model dominates the reduced model for every FPR / TPR pair.

```{r}
pacman::p_load(pracma)
-trapz(performance_metrics_oos_reduced_mod$FPR, performance_metrics_oos_reduced_mod$recall)
-trapz(performance_metrics_oos_full_mod$FPR, performance_metrics_oos_full_mod$recall)
```

As we lose information that is related to the true causal inputs, we lose predictive ability. Same story for this entire data science class since error due to ignorance increases! And certainly no different in probabilistic classifiers.

Here's the same story with the DET curve:

```{r}
ggplot(rbind(
  performance_metrics_oos_full_mod[, model := "full"],
  performance_metrics_oos_reduced_mod[, model := "reduced"]
)) +
  geom_point(aes(x = FDR, y = miss_rate, shape = model, col = p_th), size = 1) +
  geom_abline(intercept = 0, slope = 1) + 
  coord_fixed() + xlim(0, 1) + ylim(0, 1) + 
  scale_colour_gradientn(colours = rainbow(5))
```


# Choosing a Decision Threshold Based on Asymmetric Costs and Rewards

The ROC and DET curves gave you a glimpse into all the possible classification models derived from a probability estimation model. Each point on that curve is a separate $g(x)$ with its own performance metrics. How do you pick one?

Let's create rewards and costs. Imagine we are trying to predict income because we want to sell people an expensive item e.g. a car. We want to advertise our cars via a nice packet in the mail. The packet costs \$5. If we send a packet to someone who really does make $>50K$/yr then we are expected to make \$1000. So we have rewards and costs below:

```{r}
r_tp = 0
c_fp = -5
c_fn = -1000
r_tn = 0
```

Let's return to the linear logistic model with all features. Let's calculate the overall oos average reward per observation (per person) for each possible $p_{th}$:

```{r}
n = nrow(adult_test)
performance_metrics_oos$avg_reward = 
  (r_tp * performance_metrics_oos$TP +
  c_fp * performance_metrics_oos$FP +
  c_fn * performance_metrics_oos$FN +
  r_tn * performance_metrics_oos$TN) / n
```

Let's plot average reward (reward per person) by threshold:

```{r}
ggplot(performance_metrics_oos) +
  geom_line(aes(x = p_th, y = avg_reward))
```

Obviously, the best decision is $p_{th} \approx 0$ which means you classifiy almost everything as a positive. This makes sense because the mailing is so cheap. What are the performance characteristics of the optimal model?

```{r}
i_star = which.max(performance_metrics_oos$avg_reward)
performance_metrics_oos[i_star, ]
```

The more interesting problem is where the cost of advertising is higher:

```{r}
r_tp = 0
c_fp = -200
c_fn = -1000
r_tn = 0
performance_metrics_oos$avg_reward = 
  (r_tp * performance_metrics_oos$TP +
  c_fp * performance_metrics_oos$FP +
  c_fn * performance_metrics_oos$FN +
  r_tn * performance_metrics_oos$TN) / n
ggplot(performance_metrics_oos) +
  geom_point(aes(x = p_th, y = avg_reward), lwd = 0.01)
```

What are the performance characteristics of the optimal model?

```{r}
i_star = which.max(performance_metrics_oos$avg_reward)
performance_metrics_oos[i_star, ]
```

If $g_{pr}$ is closer to $f_{pr}$, what happens? 

All the threshold-derived classification models get better and you are guaranteed to make more money since you have a better discriminating eye.

There is also a way to make asymmetric models with trees. We will do this later to give you time for this to sink in. But first...


# Bias-Variance Decomposition of Generalization Error


Let's try to fit a quadratic $f$ with a linear model and examine bias-variance tradeoff.

```{r}
rm(list = ls())
xmin = 0
xmax = 5
n_train = 20
n_test = 1000
sigsq = 1
resolution = 10000

Nsim = 1000

training_gs = matrix(NA, nrow = Nsim, ncol = 2)
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)
x_trains = matrix(NA, nrow = Nsim, ncol = n_train)
y_trains = matrix(NA, nrow = Nsim, ncol = n_train)
for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_train = x_train^2 + delta_train
  x_trains[nsim, ] = x_train
  y_trains[nsim, ] = y_train
  
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ ., data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigsq) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = x_test^2 + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Graph a few:

```{r}
pacman::p_load(ggplot2)
ggplot() + 
  geom_point(aes(x, y), data = data.frame(x = x_trains[1, ], y = y_trains[1, ]), col = "blue") +
  geom_point(aes(x, y), data = data.frame(x = x_trains[2, ], y = y_trains[2, ]), col = "darkgreen") +
  geom_point(aes(x, y), data = data.frame(x = x_trains[3, ], y = y_trains[3, ]), col = "red")
```

The blue dataset is one possible $\mathbb{D}$, the green dataset is another possible $\mathbb{D}$ and the red dataset is another possible $\mathbb{D}$. This illustrated "dataset-dataset variability".

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

This should be irreducible error plus bias-squared plus variance.

Let's visualize the bias

```{r}
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
ggplot(data.frame(x = x, f = f(x)), aes(x, f)) + 
  geom_line(col = "darkgreen") + 
  geom_abline(intercept = g_average[1], slope = g_average[2], col = "red")
```

What is the average bias of $g$?

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot() + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 100)){ #otherwise takes too long
  plot_obj = plot_obj + geom_abline(intercept = training_gs[nsim, 1], slope = training_gs[nsim, 2], col = "blue")
}

plot_obj +
  geom_abline(intercept = g_average[1], slope = g_average[2], col = "red", lwd = 2)
```

Now calculate this average variance:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```

This is not exactly equal due to numerical error.

Let's try the whole thing again using a quadratic regression!

```{r}
training_gs = matrix(NA, nrow = Nsim, ncol = 3)
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_train = x_train^2 + delta_train
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ poly(x, 2, raw = TRUE), data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigsq) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = x_test^2 + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Much lower! Why? Bias went down. 

Let's visualize the bias

```{r}
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
ggplot(data.frame(x = x, f = f(x)), aes(x, f)) + 
  geom_line(col = "darkgreen") + 
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2}, col = "red")
```

Not much! What is the average bias of $g$?

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x + g_average[3] * x^2
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2), mapping = aes(x, y), col = "blue")
}

plot_obj +
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2}, col = "red", lwd = 2)
```

Now calculate this average variance:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x + g_average[3] * x^2

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```


Try it again with quintic polynomials!

```{r}
training_gs = matrix(NA, nrow = Nsim, ncol = 6)
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_train = x_train^2 + delta_train
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ poly(x, 5, raw = TRUE), data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigsq) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = x_test^2 + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Much higher! Why? Variance went up!

Let's visualize the bias

```{r}
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
ggplot(data.frame(x = x, f = f(x)), aes(x, f)) + 
  geom_line(col = "darkgreen") + 
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5}, col = "red")
```

Not much! Now acutllay compute the average bias squared of $g$:

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

This appears to have increased over last time ... but it's only because we're not running the regression infinite times. Remember this "expectation" is only an average.

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2 + training_gs[nsim, 4] * x^3 + training_gs[nsim, 5] * x^4 + training_gs[nsim, 6] * x^5), mapping = aes(x, y), col = "blue")
}

plot_obj +
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5}, col = "red", lwd = 2)
```

It looks awful!!!

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2 + training_gs[nsim, 4] * x^3 + training_gs[nsim, 5] * x^4 + training_gs[nsim, 6] * x^5
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```

# Bias - Variance Decomposition of MSE in Regression Trees

Let's return to the simulated sine curve data which we used to introduce regression trees.

```{r}
rm(list = ls())
n_train = 100
n_test = 500
xmin = 0
xmax = 10
sigsq = 0.09
Nsim = 250
resolution = 1000
```

And load the tree package:

```{r}
options(java.parameters = "-Xmx4000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
pacman::p_load(YARF, tidyverse)
```

Now let's generate lots of different datasets and fit many tree models. Note there's a new argument `calculate_oob_error = FALSE`. This is here for speed only. Ignore this for now as we will go over what this means later in detail.

```{r}
training_gs = list() #storing entire objects - need a hash
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq))
  y_train = sin(x_train) + delta_train
  # ggplot(data.frame(x = x, y = y), aes(x, y)) + geom_point(lwd=0.6) 
  
  #fit a model g | x's, delta's and save it
  g_model = YARFCART(data.frame(x = x_train), y_train, calculate_oob_error = FALSE)
  training_gs[[nsim]] = g_model
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = sin(x_test) + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Let's look at the last data set to remind ourselves of the problem setting:

```{r}
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```


What does the storage of all the models look like?

```{r}
head(training_gs)
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Let's visualize the bias

```{r}
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#now estimate the expectation of g by averaging all the different models
g_avg_x = array(0, resolution)
for (nsim in 1 : Nsim){
  g_nsim = training_gs[[nsim]]
  g_avg_x = g_avg_x + predict(g_nsim, data.frame(x = x))
}
g_avg_x = g_avg_x / Nsim #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, expe_g = g_avg_x)) + 
  geom_line(aes(x, f), col = "darkgreen") + 
  geom_line(aes(x, expe_g), col = "red")
```

Not much! Now actually compute the average bias squared of $g$:

```{r}
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

This is small - why??

It's because trees are so expressive and have such model complexity that they can nail almost any true $f$ function!

That means the MSE save the irreducible noise is coming from the variance. Let's look at one of the lines:

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) #+ ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 1)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = g_x), aes(x = x, y = y), col = "blue")
}

plot_obj +
  geom_line(data = data.frame(x = x, expe_g = g_avg_x), mapping = aes(x, expe_g), col = "red", lwd = 1.5)
```

It looks awful!!!

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

var_x_s = array(NA, min(Nsim, 50))
for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  var_x_s[nsim] = mean((g_x - g_avg_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```

# Model Averaging via Bootstrap Aggregation ("Bagging")

```{r}
rm(list = ls())
n_train = 100
n_test = 500
xmin = 0
xmax = 10
sigsq = 0.09
num_trees = 250
resolution = 1000
```

And load the tree package:

```{r}
options(java.parameters = "-Xmx4000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
pacman::p_load(YARF)
```

Now let's generate one dataset (only ONE) 

```{r}
x_train = runif(n_train, xmin, xmax)
delta_train = rnorm(n_train, 0, sqrt(sigsq))
y_train = sin(x_train) + delta_train
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```

Spend a moment to appreciate that this is all we're going to have. There is a lot of irreducible noise that is going to prevent us from finding the sine curve.

Now we create many bootstrap samples to create similiar but different datasets and fit many tree models:

```{r}
bootstrapped_gs = list() #storing entire objects - need a hash

for (t in 1 : num_trees){
  #fit a model g | x's, delta's and save it
  bootstrap_indices_t = sample(1 : n_train, replace = TRUE)
  g_model = YARFCART(data.frame(x = x_train[bootstrap_indices_t]), y_train[bootstrap_indices_t], calculate_oob_error = FALSE)
  bootstrapped_gs[[t]] = g_model
}
```

Now let's aggregate all models constructed with bootstrap samples together by averaging and see what it looks like relative to real $f$:

```{r}
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#create the bagged model predictions
g_bagged = array(0, resolution)
for (t in 1 : num_trees){
  g_t = bootstrapped_gs[[t]]
  g_bagged = g_bagged + predict(g_t, data.frame(x = x))
}
g_bagged = g_bagged / num_trees #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, g_bagged = g_bagged)) + 
  geom_line(aes(x, f), col = "darkgreen") + 
  geom_line(aes(x, g_bagged), col = "blue")
```
  
That's pretty good considering the plot of the raw data from above.

Now let's see if this is truly better than one tree. The way to do this is generate lots of datasets and try to estimate mse (with bias and variance).

To do this, we employ a convenience method called `YARFBAG`. We first define a number of trees in the "bag":

```{r}
num_trees = 250 #called "M" in class
```

And then we run the same simulation as we did above with one tree.

```{r}
Nsim = 250
training_gs = list() #storing entire objects - need a hash
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq))
  y_train = sin(x_train) + delta_train
  # ggplot(data.frame(x = x, y = y), aes(x, y)) + geom_point(lwd=0.6) 
  
  #fit a model g | x's, delta's and save it
  g_model = YARFBAG(data.frame(x = x_train), y_train, num_trees = num_trees, calculate_oob_error = FALSE)
  training_gs[[nsim]] = g_model
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = sin(x_test) + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Let's look at the last data set to remind ourselves of the problem setting:

```{r}
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```


What does the storage of all the models look like?

```{r}
head(training_gs)
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Let's visualize the bias

```{r}
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#now estimate the expectation of g by averaging all the different models
g_avg_x = array(0, resolution)
for (nsim in 1 : Nsim){
  g_nsim = training_gs[[nsim]]
  g_avg_x = g_avg_x + predict(g_nsim, data.frame(x = x))
}
g_avg_x = g_avg_x / Nsim #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, expe_g = g_avg_x)) + 
  geom_line(aes(x, f), col = "darkgreen") + 
  geom_line(aes(x, expe_g), col = "red")
```

Not much! If anything it should be *better* than the bias of a single tree!

Now actually compute the average bias squared of $g$:

```{r}
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

Almost zilch.

Now... variance?

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) #+ ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = g_x), aes(x = x, y = y), col = "blue")
}

plot_obj +
  geom_line(data = data.frame(x = x, expe_g = g_avg_x), mapping = aes(x, expe_g), col = "red", lwd = 1.5)
```

There is variance, but less than previously when only one tree was used.

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

var_x_s = array(NA, min(Nsim, 50))
for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  var_x_s[nsim] = mean((g_x - g_avg_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```

We have a better algorithm!

# Bagged Trees vs. a Linear Model

First we'll load the packages and data:

```{r}
rm(list = ls())
options(java.parameters = "-Xmx8000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
# pacman::p_load(YARF, ggplot2)
library(YARF)
diamonds_sample = diamonds[sample(1 : nrow(diamonds), 2000), ]
boston = MASS::Boston
cancer = MASS::biopsy
cancer$ID = NULL
cancer = na.omit(cancer)
pacman::p_load_gh("coatless/ucidata")
data(adult)
adult = na.omit(adult) #kill any observations with missingness
adult_sample = adult[sample(1 : nrow(adult), 2000), ]
```

Let's look at the boston housing data first

```{r}
prop_test = 0.1
test_indices = sample(1 : nrow(boston), round(prop_test * nrow(boston)))
boston_test = boston[test_indices, ]
y_test = boston_test$medv
X_test = boston_test
X_test$medv = NULL
train_indices = setdiff(1 : nrow(boston), test_indices)
boston_train = boston[train_indices, ]
y_train = boston_train$medv
X_train = boston_train
X_train$medv = NULL

mod_lin = lm(y_train ~ ., X_train)
y_hat_test_lin = predict(mod_lin, X_test)
s_e_lin = sd(y_test - y_hat_test_lin)
s_e_lin

num_trees = 500
mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
y_hat_test_bag = predict(mod_bag, X_test)
s_e_bag = sd(y_test - y_hat_test_bag)
s_e_bag

cat("oos standard error reduction:", (1 - s_e_bag / s_e_lin) * 100, "%\n")
```

How much room was there to improve?

```{r}
summary(mod_lin)$r.squared
```

Whole lot of room to improve! That extra 26% or so of unexplained variance is split between bias and irreducible error due to unknown information. The bagged trees can get rid of the bias and minimize variance while doing so. And it did a great job!

Now the diamonds data:

```{r}
n_train = 500

training_indices = sample(1 : nrow(diamonds), n_train)
diamonds_train = diamonds[training_indices, ]
y_train = diamonds_train$price
X_train = diamonds_train
X_train$price = NULL


test_indices = setdiff(1 : nrow(diamonds), training_indices)
diamonds_test = diamonds[test_indices, ]
y_test = diamonds_test$price
X_test = diamonds_test
X_test$price = NULL


mod_lin = lm(y_train ~ ., X_train)
y_hat_test_lin = predict(mod_lin, X_test)
s_e_lin = sd(y_test - y_hat_test_lin)
s_e_lin

mod_bag = YARFBAG(X_train, y_train, calculate_oob_error = FALSE)
y_hat_test_bag = predict(mod_bag, X_test)
s_e_bag = sd(y_test - y_hat_test_bag)
s_e_bag

cat("oos standard error reduction:", (1 - s_e_bag / s_e_lin) * 100, "%\n")
```

```{r}
summary(mod_lin)$r.squared
```

Not a whole lot of room to improve! That extra 8% of unexplained variance is split between bias and irreducible error due to unknown information. But it did the best it can. It is likely what's remaining is due to information we are not privy to.


# Validation in Bagging?

We are using the "bootstrap" to get the trees. Can we do model validation in the same step? 

The answer is yes. For every tree, there was a bootstrap sample of the training set used to build the tree. But there are observations in $\mathbb{D}$ that are not in the bootstrap sample! About 1/3 on average are left out i.e. "out of bag (oob)". Over many trees, there are different oob subsets than become the full data set. So you actually have validation in a way on the whole dataset kind of like K-fold cross validation. Supposedly this validation is similar to K=2 in terms of performance. It is what everyone seems to use. 

Let's load the data and packages from last class and plot the data:

```{r}
rm(list = ls())
n_train = 100
n_test = 500
xmin = 0
xmax = 10
sigsq = 0.09
num_trees = 500
x_train = runif(n_train, xmin, xmax)
delta_train = rnorm(n_train, 0, sqrt(sigsq))
y_train = sin(x_train) + delta_train
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```

Let's look at one bagged tree model and compute OOB errors after construction:

```{r}
bagged_tree_mod = YARFBAG(data.frame(x = x_train), y_train, num_trees = num_trees) #TRUE is the default! Why? It's something you want to know!
bagged_tree_mod
```

How did this work? Let's look at the oob sets:

```{r}
cat("bootstrap indices:")
sort(bagged_tree_mod$bootstrap_indices[[1]])
cat("oob:")
sort(setdiff(1 : n_train, bagged_tree_mod$bootstrap_indices[[1]]))
cat("bootstrap indices:")
bagged_tree_mod$bootstrap_indices[[2]]
cat("oob:")
setdiff(1 : n_train, bagged_tree_mod$bootstrap_indices[[2]])
```

It took predictions on each tree on the oob set, averaged by observation across trees and then averaged across observation averages.


# Random Forests


What do we have now? We have model selection is done within training. And training and validation are done in a single step! No more costly K-fold CV with 3 splits!

But can it get any better? YES. As you saw, the variance terms can be shrunk further the more decorrelated the trees become. We do this now by introducing randomness into the splits by choosing only a subset of the features to split on randomly. The trees are then grown as normal. Then the we model average many trees via bagging. And that's random forests!

Quick demo with the boston housing data and oob validation:

```{r}
rm(list = ls())
boston = MASS::Boston
y = boston$medv
X = boston
X$medv = NULL

mod_bag = YARFBAG(X, y, num_trees = 500)
mod_bag
mod_rf = YARF(X, y, num_trees = 500)
mod_rf
```

Gain in decorrelation?

```{r}
cat("gain: ", (mod_rf$pseudo_rsq_oob - mod_bag$pseudo_rsq_oob) / mod_bag$pseudo_rsq_oob * 100, "%\n")
```

For this example, not much. How about on the diamonds dataset?

```{r}
diamonds = ggplot2::diamonds

rm(list = ls())
num_trees = 500
n_train = 500

training_indices = sample(1 : nrow(diamonds), n_train)
diamonds_train = diamonds[training_indices, ]
y_train = diamonds_train$price
X_train = diamonds_train
X_train$price = NULL

mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
mod_rf = YARF(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)

test_indices = setdiff(1 : nrow(diamonds), training_indices)
diamonds_test = diamonds[test_indices, ]
y_test = diamonds_test$price
X_test = diamonds_test
X_test$price = NULL

mse_bag = sum((y_test = predict(mod_bag, X_test))^2) / nrow(X_test)
mse_rf = sum((y_test = predict(mod_rf, X_test))^2) / nrow(X_test)
cat("gain: ", (mse_bag - mse_rf) / mse_bag * 100, "%\n")
```

Also, not much, but real.


# Correlation does not Imply Causation

Take a look at the following real data:

```{r}
spurious = data.frame(
  yearly_divorce_rate_maine_per_1000 = c(5,4.7,4.6,4.4,4.3,4.1,4.2,4.2,4.2,4.1),
  yearly_US_consumption_margarine_per_capita = c(8.2,7,6.5,5.3,5.2,4,4.6,4.5,4.2,3.7)
)

ggplot(spurious, aes(x = yearly_divorce_rate_maine_per_1000, y = yearly_US_consumption_margarine_per_capita)) +
  geom_point() + geom_smooth()
```

How did this happen? 

Well, we can imagine doing the same thing.


```{r}
y = spurious$yearly_US_consumption_margarine_per_capita
n = length(y)

r = 100000
best_abs_corr = 0
best_random_xs = NULL
for (i in 1 : r){
  x = rnorm(n)
  
  random_abs_corr = abs(cor(x, y))
  if (random_abs_corr > best_abs_corr){
    best_abs_corr = random_abs_corr
    best_random_xs = x
  }
}
spurious$best_random_xs = best_random_xs
ggplot(spurious, aes(x = best_random_xs, y = yearly_US_consumption_margarine_per_capita)) +
  geom_point() + geom_smooth() + ggtitle(paste("Spurious Correlation has |r| = ", round(best_abs_corr, 3)))
```

So what's the narrative here? If you look through a gajillion random features that have no causal connection with the phenomenon $y$, you will eventually find something that "clicks". However, this will be vanish if you keep collecting data. Anything that is built upon falsehood will crumble!

#An Example of Correlation without Causation

When does correlation really not imply causation? From class, we spoke about the phenomenon $y$ = "num car accidents" with observed feature "x" = "num umbrellas sold" but common cause $z$ = "rain amount". It is clear the umbrella sales has *no causal* relationship with car accidents. But they *are correlated* because they are linked by a common cause. Here's the data example that makes this clear.

The data generating process as specified by the causal diagram looks as follows:

```{r}
rm(list = ls())
set.seed(1)
n = 300
sigma = 0.3

umbrella_example_data = data.frame(
  z_rainfall = runif(n, 0, 6) #here's the common cause - rainfall
)
umbrella_example_data$x_umbrella_sales = umbrella_example_data$z_rainfall^2 + rnorm(n, sigma) #x is a variable that is driven by z with noise
umbrella_example_data$y_car_accidents = umbrella_example_data$z_rainfall + rnorm(n, sigma) #y is a variable driven by z with noise
```

So we only see $x$ and $y$. Here's what it appears as:

```{r}
pacman::p_load(ggplot2)
ggplot(umbrella_example_data) +
  aes(x = x_umbrella_sales, y = y_car_accidents) +
  geom_point() + 
  geom_smooth(method = "lm")
```

and the model looks like:

```{r}
mod = lm(y_car_accidents ~ x_umbrella_sales, umbrella_example_data)
round(coef(mod), 3)
```

So what's the interpretation of the coefficient for $x$? ...

What you can't say is that $x$ is a causal contributor to $y$! You may want to say it, but you can't!

Now let's build a model of $y$ linear in both $x$ and $z$. What happens?

```{r}
mod = lm(y_car_accidents ~ x_umbrella_sales + z_rainfall, umbrella_example_data)
round(coef(mod), 3)
```

The effect of $x$ is gone!! Why? If you keep $z$ constant, the sole true causal factor in $y$, manipulating $x$ won't matter anymore!

Why is this? Well, you can look at how x affects y in local areas of z for instance.

```{r}
b = 0.8; a = 0.7
z_small_indices = umbrella_example_data$z_rainfall < 
  quantile(umbrella_example_data$z_rainfall, b) &
  umbrella_example_data$z_rainfall >
  quantile(umbrella_example_data$z_rainfall, a)

ggplot(umbrella_example_data[z_small_indices, ]) +
  aes(x = x_umbrella_sales, y = y_car_accidents) +
  geom_point()
```

If you force the common cause (lurking variable) to be an approximate constant, then you won't see any affect of x on y.


#Classification Trees and Confusion Tables

Let's load up the adult dataset where the response is 1 if the person makes more than $50K per year and 0 if they make less than $50K per year.

```{r}
pacman::p_load_gh("coatless/ucidata")
data(adult)
adult %<>% 
  na.omit #kill any observations with missingness
str(adult)
?adult
```

Let's use samples of 2,000 to run experiments:

```{r}
test_size = 2000
train_indices = sample(1 : nrow(adult), test_size)
adult_train = adult[train_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL
test_indices = sample(setdiff(1 : nrow(adult), train_indices), test_size)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL
```

Load the tree package:

```{r}
options(java.parameters = "-Xmx4000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
pacman::p_load(YARF)
```

Make a tree:

```{r}
tree_mod = YARFCART(X_train, y_train)
```

How "big" is this tree model?

```{r}
get_tree_num_nodes_leaves_max_depths(tree_mod)
```

What are the "main" splits?

```{r}
illustrate_trees(tree_mod, max_depth = 4, open_file = TRUE)
```

Compute in-sample and out of sample fits:

```{r}
y_hat_train = predict(tree_mod, X_train)
y_hat_test = predict(tree_mod, X_test)
```

Let's look at the confusion table in-sample:

```{r}
table(y_train, y_hat_train)
```

There are no errors here! Thus, precision and recall are both 100%. This makes sense because classification trees overfit.

Let's do the same oos:

```{r}
oos_conf_table = table(y_test, y_hat_test)
oos_conf_table
```

We didn't do as well (of course). Let's calculate some performance metrics. We assume ">50k" is the "positive" category and "<=50k" is the "negative" category. Note that this choice is arbitrary and everything would just be switched if we did it the other way.

```{r}
n = sum(oos_conf_table)
n
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])
acc = (tp + tn) / n
acc
misclassifcation_error = 1 - acc
misclassifcation_error
precision = tp / num_pred_pos
precision
recall = tp / num_pos
recall
false_discovery_rate = 1 - precision
false_discovery_rate
false_omission_rate = fn / num_pred_neg
false_omission_rate
```

Let's see how this works on a dataset whose goal is classification for more than 2 levels. Note: this is only possible now with trees!

```{r}
rm(list = ls())
pacman::p_load(mlbench)
data(LetterRecognition)
LetterRecognition = na.omit(LetterRecognition) #kill any observations with missingness
str(LetterRecognition)
?LetterRecognition
```

Now we split the data:

```{r}
test_samp = 2000
train_indices = sample(1 : nrow(LetterRecognition), test_samp)
ltr_train = LetterRecognition[train_indices, ]
y_train = ltr_train$lettr
X_train = ltr_train
X_train$lettr = NULL
test_indices = sample(setdiff(1 : nrow(LetterRecognition), train_indices), test_samp)
ltr_test = LetterRecognition[test_indices, ]
y_test = ltr_test$lettr
X_test = ltr_test
X_test$lettr = NULL
```

And fit a tree model and its in-sample and oos fits:

```{r}
tree_mod = YARFCART(X_train, y_train)
y_hat_train = predict(tree_mod, X_train)
y_hat_test = predict(tree_mod, X_test)
```

Take a look at the in-sample confusion matrix:

```{r}
table(y_train, y_hat_train)
```

Perfecto... as expected... 

Now the oos confusion matrix:

```{r}
oos_confusion_table = table(y_test, y_hat_test)
oos_confusion_table
```

Hard to read. Let's make it easier to read by blanking out the diagonal and looking at entried only >= 5:

```{r}
oos_confusion_table[oos_confusion_table < 5] = ""
diag(oos_confusion_table) = "."
oos_confusion_table
```

What's it using to determine letter?

```{r}
illustrate_trees(tree_mod, max_depth = 3, open_file = TRUE)
```

Where did these features comes from?? Deep learning helps to create the features from the raw pixel data. Wish I had a whole next semester to discuss this...

Random Forests:

```{r}
num_trees = 500
n_train = 500

training_indices = sample(1 : nrow(adult), n_train)
adult_train = adult[training_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL

mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
mod_rf = YARF(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
```

And test:

```{r}
test_indices = sample(setdiff(1 : nrow(adult), training_indices), 25000)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL

y_hat_test_bag = predict(mod_bag, X_test)
y_hat_test_rf = predict(mod_rf, X_test)

oos_conf_table_bag = table(y_test, y_hat_test_bag)
oos_conf_table_rf = table(y_test, y_hat_test_rf)
oos_conf_table_bag
oos_conf_table_rf
miscl_err_bag = mean(y_test != y_hat_test_bag)
miscl_err_rf = mean(y_test != y_hat_test_rf)
miscl_err_bag
miscl_err_rf

cat("gain: ", (miscl_err_bag - miscl_err_rf) / miscl_err_bag * 100, "%\n")
```

And on letters:

```{r}
test_samp = 2000
train_indices = sample(1 : nrow(LetterRecognition), test_samp)
ltr_train = LetterRecognition[train_indices, ]
y_train = ltr_train$lettr
X_train = ltr_train
X_train$lettr = NULL
test_indices = sample(setdiff(1 : nrow(LetterRecognition), train_indices), test_samp)
ltr_test = LetterRecognition[test_indices, ]
y_test = ltr_test$lettr
X_test = ltr_test
X_test$lettr = NULL
```

And fit a tree model and its in-sample and oos fits:

```{r}
mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
mod_rf = YARF(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
mod_bag
mod_rf

oos_conf_table_bag = table(y_test, y_hat_test_bag)
oos_conf_table_rf = table(y_test, y_hat_test_rf)
oos_conf_table_bag
oos_conf_table_rf
miscl_err_bag = mean(y_test != y_hat_test_bag)
miscl_err_rf = mean(y_test != y_hat_test_rf)
miscl_err_bag
miscl_err_rf

cat("gain: ", (miscl_err_bag - miscl_err_rf) / miscl_err_bag * 100, "%\n")
```


# Missingness

Take a look at the weather dataset again:

```{r}
pacman::p_load(nycflights13, tidyverse, magrittr)
data(weather)
str(weather)
summary(weather)
```

Note how `time_hour` is perfectly collinear with the year, month, day and hour columns. So let's drop `time_hour`:

```{r}
weather %<>%
  select(-c(time_hour))
```

We also cannot have character variables:

```{r}
weather %<>%
  mutate(origin = as.factor(origin))
```


Imagine we were trying to predict `precip`. So let's section our dataset:

```{r}
y = weather$precip
X = weather
X$precip = NULL
rm(weather)
```

Let's first create a matrix with $p$ columns that represents missingness

```{r}
M = tbl_df(apply(is.na(X), 2, as.numeric))
colnames(M) = paste("is_missing_", colnames(X), sep = "")
head(M)
summary(M)
```

Some of these missing indicators are collinear because they share all the rows they are missing on. Let's filter those out:

```{r}
M = tbl_df(t(unique(t(M))))
```


Many featuers did not have missingness so let's remove them:

```{r}
M %<>% select_if(function(x){sum(x) > 0})
head(M)
dim(M)
colSums(M)
```

Now let's work on imputation. The missingness is not extreme except in the `wind_gust` variable. Let's drop that variable:

```{r}
X %<>% select(-wind_gust)
```

Without imputing and without using missingness as a predictor in its own right, let's see what we get with a basic linear model now:

```{r}
lin_mod_listwise_deletion = lm(y ~ ., X)
summary(lin_mod_listwise_deletion)
```

A measly 13.2%. Also note: year is collinear.

Now let's impute using the package. we cannot fit RF models to the entire dataset (it's 26,000! observations) so we will sample 2,000 for each of the trees and then average. That will be good enough.

```{r}
pacman::p_load(missForest)
Ximp = missForest(data.frame(X), sampsize = rep(2000, ncol(X)))$ximp
```

Now we take our imputed dataset, combine it with our missingness indicators for a new design matrix.

```{r}
Xnew = data.frame(cbind(Ximp, M))
linear_mod_impute_and_missing_dummies = lm(y ~ ., Xnew)
summary(linear_mod_impute_and_missing_dummies)
```

Is this a better model?? Are they even comparable? How to compare them? I'm not sure... certainly an R^2 comparison is the wrong way to compare.

Note: this is just an illustration of best practice. It didn't necessarily have to "work".

It is hard to compare the two models since the first model was built with 23,000 observations and this was built with the full 26,000 observations. Those extra 3,000 are the most difficult to predict on. This is complicated...